---
title: "Data preparation notebook"
output: html_notebook
---

Author: lprtk

Data Mining for credit risk


# Pre-processing et data preparation


#### Librairies
```{r}
# Define the libraries to be used
libraries_used <- 
  c("arules", "caret","corrplot" ,"dplyr", "lubridate", "mice", 
    "naniar", "stringr", "superml", "tidyr", "UpSetR", "visdat", "VIM")

# Verification of installed libraries
libraries_missing <- 
  libraries_used[!(libraries_used %in% installed.packages()[,"Package"])]

# Install missing libraries
if(length(libraries_missing)) install.packages(libraries_missing)
```


#### Librairies import
```{r}
library(arules)
library(caret)
library(corrplot)
library(dplyr)
library(lubridate)
library(mice)
library(naniar)
library(stringr)
library(superml)
library(tidyr)
library(UpSetR)
library(visdat)
library(VIM)
```

-------------------------------------------------------------------------------------------------

#### Function definition
```{r}
# Function to discretize a variable
discretizer <- function(column) {
  
  column <- cut(column, breaks = c(0, 12, 24, 60, 120, Inf) , labels = c(1, 2, 3, 4, 5),
                include.lowest = FALSE, right = TRUE)
}

# Function to transform a character variable into a factor
to_factor <- function(variable) {
  
  variable <- as.factor(variable)
}
```

-------------------------------------------------------------------------------------------------

#### Data import
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/datamining")
```


```{r}
data <- read.csv("data_lending_club.csv")
```


#### Data description
```{r}
str(data)
```


#### Data header
```{r}
head(data)
```


#### Split Test Train of the database
```{r}
trainDT <- sample(1:nrow(data), 0.8*nrow(data))
train <- data[trainDT,]
test <- data[-trainDT,]
```

-------------------------------------------------------------------------------------------------

# First cleaning of the database

`
- Deletion of variables called "useless" because they have a unique value per client, they are non-disciminates or they provide the same information as other variables;
- Deletion of the variables called "unavailable" because we do not have this information at the time of the study of the client's file to grant him a possible loan.
`
#### Removal of certain features
```{r}
# Unnecessary columns: do not allow to discriminate a client, or redundant information
useless <- c("id", "member_id", "url", "zip_code", "title", "desc", "pymnt_plan", "emp_title", "initial_list_status", "policy_code", "addr_state", "sub_grade", "issue_d", "funded_amnt", "funded_amnt_inv", "verification_status_joint", "disbursement_method", "earliest_cr_line")

# Variables unavailable at the time of the client's file review
unavailable <- c("out_prncp", "out_prncp_inv", "total_pymnt", "total_pymnt_inv", "total_rec_prncp", "funded_amnt_inv", "total_rec_int", "total_rec_late_fee", "recoveries", "collection_recovery_fee", "next_pymnt_d", "last_credit_pull_d", "last_pymnt_d", "last_pymnt_amnt")

data <- data[,!(names(data) %in% useless)]
data <- data[,!(names(data) %in% unavailable)]
```


#### Visualization of missing values
```{r}
vis_miss(data, warn_large_data = FALSE)
```

#### Supression of variables with 100% missing values
```{r}
# Variables with 100% missing values
missing_values <- sapply(data, function(x) sum(is.na(x))) 
missing_values[missing_values == nrow(data)]
```

```{r}
all_na = c("revol_bal_joint", "sec_app_fico_range_high", "sec_app_inq_last_6mths", "sec_app_open_acc", "sec_app_open_act_il", "sec_app_chargeoff_within_12_mths", "sec_app_mths_since_last_major_derog", "sec_app_fico_range_low", "sec_app_earliest_cr_line", "sec_app_mort_acc", "sec_app_revol_util", "sec_app_num_rev_accts", "sec_app_collections_12_mths_ex_med")

data <- data[,!(names(data) %in% all_na)]
```


#### Definition of the target variable
```{r}
data <- data[!(data$loan_status == "Current" | data$loan_status == "In Grace Period"),]

data$default <- ifelse(data$loan_status == "Fully Paid"| data$loan_status == "Does not meet the credit policy. Status:Fully Paid", "no", "yes")
```

#### Verification
```{r}
table(data$default)
```

-------------------------------------------------------------------------------------------------

# Feature engeneering

`
- Discretization;
- Type transformation and format;
- Union of some variables;
- New deletion of variables.
`

#### Calls the "discretizer" function for several features
```{r}
discretizer(data$mths_since_last_delinq)
discretizer(data$mths_since_last_record)
discretizer(data$mths_since_last_major_derog)
discretizer(data$mths_since_recent_revol_delinq)
discretizer(data$mths_since_recent_bc_dlq)
discretizer(data$mths_since_rcnt_il)
discretizer(data$mo_sin_old_il_acct)
discretizer(data$mo_sin_old_rev_tl_op)
discretizer(data$mo_sin_rcnt_rev_tl_op)
discretizer(data$mo_sin_rcnt_tl)
discretizer(data$mths_since_recent_bc)
discretizer(data$mths_since_recent_inq)

# Replacement of NA by 0 for all discretized variables because they are not "true" missing values
data <- data %>% mutate( 
  mths_since_last_delinq = ifelse(is.na(mths_since_last_delinq), 0, mths_since_last_delinq),
  mths_since_last_record = ifelse(is.na(mths_since_last_record), 0, mths_since_last_record),
  mths_since_last_major_derog = ifelse(is.na(mths_since_last_major_derog), 0, mths_since_last_major_derog),
  mths_since_recent_revol_delinq = ifelse(is.na(mths_since_recent_revol_delinq), 0, mths_since_recent_revol_delinq),
  mths_since_recent_bc_dlq = ifelse(is.na(mths_since_recent_bc_dlq), 0, mths_since_recent_bc_dlq),
  mths_since_rcnt_il = ifelse(is.na(mths_since_rcnt_il), 0, mths_since_rcnt_il),
  mo_sin_old_il_acct = ifelse(is.na(mo_sin_old_il_acct), 0, mo_sin_old_il_acct),
  mo_sin_old_rev_tl_op = ifelse(is.na(mo_sin_old_rev_tl_op), 0, mo_sin_old_rev_tl_op),
  mo_sin_rcnt_rev_tl_op = ifelse(is.na(mo_sin_rcnt_rev_tl_op), 0, mo_sin_rcnt_rev_tl_op),
  mo_sin_rcnt_tl = ifelse(is.na(mo_sin_rcnt_tl), 0, mo_sin_rcnt_tl),
  mths_since_recent_bc = ifelse(is.na(mths_since_recent_bc), 0, mths_since_recent_bc),
  mths_since_recent_inq = ifelse(is.na(mths_since_recent_inq), 0, mths_since_recent_inq)
  )
```


#### Transform character to factor
```{r}
to_factor(data$grade)
to_factor(data$verification_status)
to_factor(data$home_ownership)
to_factor(data$purpose)
to_factor(data$application_type)
```


#### Conversion of the "term" variable into a numerical variable
```{r}
data$term <- str_replace_all(data$term, " months", "")
data$term <- as.numeric(data$term)
```


#### Conversion of the variable "emp length" into a numerical variable
```{r}
data$emp_length <- str_replace_all(data$emp_length, " years","") 
data$emp_length <- str_replace_all(data$emp_length, " year","")
data$emp_length <- str_replace_all(data$emp_length, "< 1","0")
data$emp_length <- str_replace_all(data$emp_length, "[+]","") 

data$emp_length <- as.numeric(data$emp_length)
```


#### Grouping of joint borrowers into a single column
```{r}
data$annual_inc <- pmax(data$annual_inc_joint, data$annual_inc, na.rm = TRUE)
data$dti <- pmin(data$dti, data$dti_joint, na.rm = TRUE)

```


#### Deletion of certain variables following feature engineering
```{r}
second_drop <- c("annual_inc_joint", "dti_joint", "loan_status")
    
hardships <- c("orig_projected_additional_accrued_interest", "hardship_end_date", "hardship_start_date", "hardship_type", "hardship_reason", "hardship_status", "deferral_term", "hardship_last_payment_amount", "hardship_payoff_balance_amount", "hardship_loan_status", "hardship_dpd", "hardship_length", "payment_plan_start_date", "hardship_amount", "hardship_flag")
    
settlement <- c("settlement_term", "debt_settlement_flag_date", "settlement_status", "settlement_date","settlement_amount", "settlement_percentage", "debt_settlement_flag")

data <- data[,!(names(data) %in% second_drop)]
data <- data[,!(names(data) %in% hardships)]
data <- data[,!(names(data) %in% settlement)]
```

--------------------------------------------------------------------------------------------------

#### Outliers management
```{r}

data <- filter(data, annual_inc < quantile(data$annual_inc, probs = 0.95))
data <- filter(data, int_rate < quantile(data$int_rate, probs = 0.95))
```

--------------------------------------------------------------------------------------------------

# Treatment of missing values 

`
- Detection of missing values;
- Identification of the process;
- Imputation according to the nature of the missing value.
`

#### Summary of remaining missing values
```{r}
missing_values  <- sapply(data, function(x) sum(is.na(x))) 
sort(missing_values[missing_values != 0])
```


## Missing at Random (MAR)

#### Treatment of missing values : MAR
```{r}
na_MAR = c("open_acc_6m", "open_act_il", "open_il_12m", "open_il_24m", "total_bal_il", "open_rv_12m", "open_rv_24m", "max_bal_bc", "all_util" , "inq_fi", "total_cu_tl", "inq_last_12m")

miss_data_MAR <- select(data,all_of(na_MAR))

gg_miss_upset(miss_data_MAR, nsets = 20)
```

#### Multiple imputation on mixed data
```{r}
imputed <- mice(miss_data_MAR, meth = "pmm", m = 5, printFlag = FALSE)

# see imputed values (5 imputations per variable)
imputed$imp
```

#### Storage of the 5 multiple imputations
```{r}
imputed_data_MAR_1 <- complete(imputed, action = 1)
imputed_data_MAR_2 <- complete(imputed, action = 2)
imputed_data_MAR_3 <- complete(imputed, action = 3)
imputed_data_MAR_4 <- complete(imputed, action = 4)
imputed_data_MAR_5 <- complete(imputed, action = 5)
```

#### Checking for missing values after imputation
```{r}
gg_miss_var(imputed_data_MAR_2)
```


## Missing Completly at Random (MCAR)

#### Treatment of missing values : MCAR
```{r}
missing_values  = sapply(data, function(x) sum(is.na(x))) 
sort(missing_values[missing_values != 306 & missing_values != 0 ])
```

```{r}
na_MCAR = c("revol_util", "bc_open_to_buy", "bc_util", "percent_bc_gt_75", "num_tl_120dpd_2m", "emp_length", "il_util")

miss_data_MCAR <- select(data,all_of(na_MCAR))
gg_miss_upset(miss_data_MCAR, nsets = 20)
```

#### Imputation using the kNN method
```{r}
imputed_data_MCAR <-  kNN(miss_data_MCAR)
data_imputed <- cbind(imputed_data_MAR_2,imputed_data_MCAR[,1:7])
```

#### Visualization of the remaining missing values
```{r}
gg_miss_var(data_imputed)
```


#### Construction of the final database
```{r}
# Retrieving the name of columns containing NA
names(data_imputed)
```

```{r}
na_var <- c("open_acc_6m", "open_act_il", "open_il_12m", "open_il_24m", "total_bal_il",    "open_rv_12m", "open_rv_24m", "max_bal_bc", "all_util", "inq_fi", "total_cu_tl", "inq_last_12m", "revol_util", "bc_open_to_buy","bc_util", "percent_bc_gt_75", "num_tl_120dpd_2m", "emp_length", "il_util")

# Creation of a table without the variables having NA (Objective: to rebuild the initial base by replacing the NA by the imputed values)
data_without_na <- data[,!(names(data) %in% na_var)]
```

#### Final dataframe
```{r}
data_cleaned <- cbind(data_without_na, data_imputed)
```


#### Check if the missing values of the database have been cleaned
```{r}
gg_miss_var(data_cleaned)
```

--------------------------------------------------------------------------------------------------

# Study of correlations


#### Calculation of correlations
```{r}
corr_simple <- function(data =  data_cleaned, sig = 0.8){
  
  df_cor <- data %>% mutate_if(is.character, as.factor)
  df_cor <- df_cor %>% mutate_if(is.factor, as.numeric)
  
  corr <- cor(df_cor)
      
  corr[lower.tri(corr,diag=TRUE)] <- NA 
  
  corr[corr == 1] <- NA 
  
  corr <- as.data.frame(as.table(corr))
   
  corr <- na.omit(corr) 
    
  corr <- subset(corr, abs(Freq) > sig) 
  
  corr <- corr[order(-abs(corr$Freq)),] 
  
  print(corr)
  
  mtx_corr <- reshape2::acast(corr, Var1~Var2, value.var="Freq")
  
  corrplot(mtx_corr, is.corr=FALSE, tl.col="black", na.label=" ")
}

corr_simple()
```

#### Suppression of overly correlated variables
```{r}
drop_corr_var <- c("fico_range_high", "num_sats", "tot_cur_bal","num_actv_rev_tl", "total_bal_il","total_bal_ex_mort","bc_open_to_buy", "percent_bc_gt_75", "revol_util", "num_bc_tl", "acc_open_past_24mths", "num_bc_sats", "num_tl_30dpd", "num_op_rev_tl", "open_rv_12m", "num_actv_rev_tl", "total_rev_hi_lim", "num_rev_tl_bal_gt_0")

data_cleaned <- data_cleaned[,!(names(data_cleaned) %in% drop_corr_var)]
```

--------------------------------------------------------------------------------------------------

# Encoding qualitative variables


#### Label Encoding
```{r}
lbl <- LabelEncoder$new()
data_cleaned$grade <- lbl$fit_transform(data_cleaned$grade)
data_cleaned$application_type <- lbl$fit_transform(data_cleaned$application_type)
```


#### One Hot Encoding
```{r}
dummy <- dummyVars(" ~ .", data = data_cleaned)

final_df <- data.frame(predict(dummy, newdata = data_cleaned))
```


#### Verification of the re-encoding on the final base
```{r}
str(final_df)
```

--------------------------------------------------------------------------------------------------

# Definition of a function containing all the pre-processing


#### Application of the pre_processing function to the train and the test
```{r}
pre_processing <- function(data) {
  
  # First deletion
  useless <- c("id", "member_id", "url", "zip_code", "title", "desc", "pymnt_plan" , "emp_title" , "initial_list_status", "policy_code", "addr_state", "sub_grade", "issue_d", "funded_amnt", "funded_amnt_inv", "verification_status_joint", "disbursement_method", "earliest_cr_line")
  
  unavailable <- c("out_prncp","out_prncp_inv","total_pymnt","total_pymnt_inv", "total_rec_prncp", "funded_amnt_inv", "total_rec_int","total_rec_late_fee", "recoveries", "collection_recovery_fee", "next_pymnt_d", "last_credit_pull_d","last_pymnt_d","last_pymnt_amnt")
  
  data <- data[,!(names(data) %in% useless)]
  data <- data[,!(names(data) %in% unavailable)]
  
  # Deletion of variables with 100% NA
  all_na <- c("revol_bal_joint", "sec_app_fico_range_high", "sec_app_inq_last_6mths", "sec_app_open_acc", "sec_app_open_act_il", "sec_app_chargeoff_within_12_mths", "sec_app_mths_since_last_major_derog", "sec_app_fico_range_low", "sec_app_earliest_cr_line", "sec_app_mort_acc", "sec_app_revol_util", "sec_app_num_rev_accts", "sec_app_collections_12_mths_ex_med")

  data <- data[,!(names(data) %in% all_na)]
  
  # Definition of the target
  data <- data[!(data$loan_status == "Current" | data$loan_status == "In Grace Period"),]
  data$default <- ifelse(data$loan_status == "Fully Paid"| data$loan_status == "Does not meet the credit policy. Status:Fully Paid", "no", "yes")
  
  # Calls the discretizer function for multiple columns
  discretizer(data$mths_since_last_delinq)
  discretizer(data$mths_since_last_record)
  discretizer(data$mths_since_last_major_derog)
  discretizer(data$mths_since_recent_revol_delinq)
  discretizer(data$mths_since_recent_bc_dlq)
  discretizer(data$mths_since_rcnt_il)
  discretizer(data$mo_sin_old_il_acct)
  discretizer(data$mo_sin_old_rev_tl_op)
  discretizer(data$mo_sin_rcnt_rev_tl_op)
  discretizer(data$mo_sin_rcnt_tl)
  discretizer(data$mths_since_recent_bc)
  discretizer(data$mths_since_recent_inq)
  
  # Replacement of NA by 0 for all discretized variables because they are not "true" missing values
  data <- data %>% mutate( 
    mths_since_last_delinq = ifelse(is.na(mths_since_last_delinq), 0, mths_since_last_delinq),
    mths_since_last_record = ifelse(is.na(mths_since_last_record), 0, mths_since_last_record),
    mths_since_last_major_derog = ifelse(is.na(mths_since_last_major_derog), 0, mths_since_last_major_derog),
    mths_since_recent_revol_delinq = ifelse(is.na(mths_since_recent_revol_delinq), 0, mths_since_recent_revol_delinq),
    mths_since_recent_bc_dlq = ifelse(is.na(mths_since_recent_bc_dlq), 0, mths_since_recent_bc_dlq),
    mths_since_rcnt_il = ifelse(is.na(mths_since_rcnt_il), 0, mths_since_rcnt_il),
    mo_sin_old_il_acct = ifelse(is.na(mo_sin_old_il_acct), 0, mo_sin_old_il_acct),
    mo_sin_old_rev_tl_op = ifelse(is.na(mo_sin_old_rev_tl_op), 0, mo_sin_old_rev_tl_op),
    mo_sin_rcnt_rev_tl_op = ifelse(is.na(mo_sin_rcnt_rev_tl_op), 0, mo_sin_rcnt_rev_tl_op),
    mo_sin_rcnt_tl = ifelse(is.na(mo_sin_rcnt_tl), 0, mo_sin_rcnt_tl),
    mths_since_recent_bc = ifelse(is.na(mths_since_recent_bc), 0, mths_since_recent_bc),
    mths_since_recent_inq = ifelse(is.na(mths_since_recent_inq), 0, mths_since_recent_inq)
    )
  
  # Transform character to factor
  to_factor(data$grade)
  to_factor(data$verification_status)
  to_factor(data$home_ownership)
  to_factor(data$purpose)
  to_factor(data$application_type)
  
  # Conversion of the variable "term" into a numerical variable
  data$term <- str_replace_all(data$term, " months", "")
  data$term <- as.numeric(data$term)
  
  # Conversion of the "emp_length" variable into a numeric variable
  data$emp_length <- str_replace_all(data$emp_length, " years", "") 
  data$emp_length <- str_replace_all(data$emp_length, " year", "")
  data$emp_length <- str_replace_all(data$emp_length, "< 1", "0")
  data$emp_length <- str_replace_all(data$emp_length, "[+]", "") 
  
  data$emp_length = as.numeric(data$emp_length)
  
  # Grouping of joint borrowers into a single column
  data$annual_inc <- pmax(data$annual_inc_joint, data$annual_inc, na.rm = TRUE)
  data$dti <- pmin(data$dti, data$dti_joint, na.rm = TRUE)
  
  # Supprimer les variables qui suivent l'ingénierie des fonctionnalités
  second_drop <- c("annual_inc_joint", "dti_joint", "loan_status")
      
  hardships <- c("orig_projected_additional_accrued_interest","hardship_end_date", "hardship_start_date", "hardship_type",   "hardship_reason","hardship_status", "deferral_term", "hardship_last_payment_amount","hardship_payoff_balance_amount","hardship_loan_status", "hardship_dpd", "hardship_length", "payment_plan_start_date", "hardship_amount", "hardship_flag")
      
  settlement <- c("settlement_term", "debt_settlement_flag_date", "settlement_status", "settlement_date","settlement_amount", "settlement_percentage", "debt_settlement_flag")
  
  data <- data[,!(names(data) %in% second_drop)]
  data <- data[,!(names(data) %in% hardships)]
  data <- data[,!(names(data) %in% settlement)]
  
  # Outliers management
  data <- filter(data, annual_inc < quantile(data$annual_inc, probs = 0.95))
  data <- filter(data, int_rate < quantile(data$int_rate, probs = 0.95))
  
  # Treatment of missing values : MAR
  na_MAR = c("open_acc_6m", "open_act_il", "open_il_12m", "open_il_24m", "total_bal_il", "open_rv_12m", "open_rv_24m", "max_bal_bc", "all_util" , "inq_fi", "total_cu_tl", "inq_last_12m")
  
  miss_data_MAR <- select(data,all_of(na_MAR))
  
  # Multiple imputation on mixed data
  imputed <- mice(miss_data_MAR, meth = "pmm", m = 5, printFlag = FALSE)
  imputed_data_MAR_2 <- complete(imputed, action = 2)

  # Treatment of missing values : MCAR
  na_MCAR <- c("revol_util", "bc_open_to_buy", "bc_util", "percent_bc_gt_75", "num_tl_120dpd_2m", "emp_length", "il_util")

  miss_data_MCAR <- select(data,all_of(na_MCAR))
  
  # Imputation using the kNN method
  imputed_data_MCAR <- kNN(miss_data_MCAR)
  data_imputed <- cbind(imputed_data_MAR_2,imputed_data_MCAR[,1:7])
  
  na_var <- c("open_acc_6m", "open_act_il", "open_il_12m", "open_il_24m", "total_bal_il",  "open_rv_12m", "open_rv_24m", "max_bal_bc", "all_util", "inq_fi", "total_cu_tl", "inq_last_12m",   "revol_util", "bc_open_to_buy","bc_util", "percent_bc_gt_75", "num_tl_120dpd_2m", "emp_length", "il_util")

  # Creation of a table without the variables having NA (Objective: to rebuild the initial base by replacing the NA with the imputed values)
  data_without_na <- data[,!(names(data) %in% na_var)]
  
  # Final dataframe
  data_cleaned = cbind(data_without_na, data_imputed)
  
  # Removal of variables that are too correlated
  drop_corr_var = c("fico_range_high", "num_sats", "tot_cur_bal","num_actv_rev_tl", "total_bal_il","total_bal_ex_mort","bc_open_to_buy", "percent_bc_gt_75", "revol_util", "num_bc_tl", "acc_open_past_24mths", "num_bc_sats", "num_tl_30dpd", "num_op_rev_tl", "open_rv_12m", "num_actv_rev_tl", "total_rev_hi_lim", "num_rev_tl_bal_gt_0")
  
  data_cleaned <- data_cleaned[,!(names(data_cleaned) %in% drop_corr_var)]
  
  # Label Encoding
  lbl = LabelEncoder$new()
  data_cleaned$grade = lbl$fit_transform(data_cleaned$grade)
  data_cleaned$application_type = lbl$fit_transform(data_cleaned$application_type)
  
  # One Hot Encoding
  dummy <- dummyVars(" ~ .", data = data_cleaned)
  final_df <- data.frame(predict(dummy, newdata = data_cleaned))
  
  return(data_cleaned)
}
```

#### Application of the pre-processing function on the trainset
```{r}
train <- pre_processing(train)
```

#### Application of the pre-processing function on the testset
```{r}
test <- pre_processing(test)
```

--------------------------------------------------------------------------------------------------

# Export

#### Export to csv
```{r}
write.csv(train, "train.csv", row.names = FALSE)
write.csv(test, "test.csv", row.names = FALSE)
```


#### Checking the target rate
```{r}
table(train$default)

table(test$default)
```

--------------------------------------------------------------------------------------------------
