---
title: "Unsupervised analysis notebook"
output: html_notebook
---

Author: lprtk

Data Mining for credit risk


# Unsupervised learning


#### Librairies
```{r}
# Define the libraries to be used
libraries_used <- 
  c("dplyr", "factoextra", "FactoMineR", "ggplot2", "PerformanceAnalytics", "plotly", "rgl", "funModeling")

# Verification of installed libraries
libraries_missing <- 
  libraries_used[!(libraries_used %in% installed.packages()[,"Package"])]

# Install missing libraries
if(length(libraries_missing)) install.packages(libraries_missing)
```


#### Librairies import
```{r}
library(dplyr)
library(factoextra)
library(FactoMineR)
library(ggplot2)
library(PerformanceAnalytics)
library(plotly)
library(rgl)
library(funModeling)
```

--------------------------------------------------------------------------------------------------

#### Data import
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/datamining")
```

```{r}
data <- read.csv("data_unsupervised.csv", sep = ",", header = T)
```


#### Data description
```{r}
detailled_data <- df_status(data, print_results = FALSE)
detailled_data
```


#### Checking the target variable
```{r}
data %>% count(loan_status)
```

--------------------------------------------------------------------------------------------------

# Principal Component Analysis (PCA)

The objective is to analyze all the variables in order to make a first dimension reduction.

#### We delete the target variable
```{r}
data_pca <- subset(data, select = -c(loan_status))
```


#### ACP initialization
```{r}
res.pca <- PCA(data_pca, scale.unit = TRUE, graph = FALSE)
```


#### Correlation circle
```{r}
fviz_pca_var(res.pca, col.var = "blue")
```


#### Eigenvalues : % of variance explained by each factorial axis
```{r}
eig.pca <- get_eigenvalue(res.pca)
```


#### Plot of inertia explained (in %) by each factorial axis
```{r}
fviz_eig(res.pca, addlabels = TRUE, ylim = c(0, 50))
```


#### Cos2 : quality of representations
```{r}
var.cos2 <- res.pca$var$cos2[,1:2]
var.cos2

var.cos2.cum <- res.pca$var$cos2[,1]+res.pca$var$cos2[,2]
var.cos2.cum

fviz_cos2(res.pca, choice = "var", axes = 1:2, top = 10)

fviz_pca_var(res.pca, col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE, title = 'Cercle de corrélation des variables')
```


#### Contribution of each variable to the axes
```{r}
var.contrib <- res.pca$var$contrib
var.contrib

fviz_pca_var(res.pca, col.var = "contrib",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"))
```


#### Contributions Plot
```{r}
fviz_contrib(res.pca, choice = "var", axes = 1, top = 30)
fviz_contrib(res.pca, choice = "var", axes = 2, top = 30)
fviz_contrib(res.pca, choice = "var", axes = 3, top = 30)
fviz_contrib(res.pca, choice = "var", axes = 4, top = 30)
```


#### Transformation of contributions into dataframe
```{r}
var.contrib <- as.data.frame(var.contrib[, c(1,2,3,4)])
```


#### Contribution filter
```{r}
var.dim1 <- filter(var.contrib, Dim.1 >= 3)
var.dim2 <- filter(var.contrib, Dim.2 >= 3)
var.dim3 <- filter(var.contrib, Dim.3 >= 3)
var.dim4 <- filter(var.contrib, Dim.4 >= 3)
```


#### Variable recovery
```{r}
col1 <- rownames(var.dim1)
col2 <- rownames(var.dim2)
col3 <- rownames(var.dim3)
col4 <- rownames(var.dim4)

col.list <- c(col1, col2, col3, col4)
```


```{r}
col.list
```

--------------------------------------------------------------------------------------------------

# Kmeans clustering

The objective is to cluster, not on the variables as we did previously with the PCA but directly on our borrowers. Can we observe similarities between borrowers who default versus others?

#### We remove the target variable
```{r}
data_kmeans <- subset(data, select = -c(loan_status))
```


#### Search for the optimal number of clusters (K)
```{r}
fviz_nbclust(data_kmeans, kmeans, method = "silhouette")
fviz_nbclust(data_kmeans, kmeans, method = "wss")
```


#### Kmeans initialization
```{r}
km <- kmeans(data_kmeans, 2)
```


#### Within Sum Square (intra-class variance: to be minimized)
```{r}
km$withinss
```


#### Between Sum Square (inter-class variance: to be maximized)
```{r}
km$betweenss
```


#### Ratio Between Sum Square / Total Sum Square: to be maximized
```{r}
km$betweenss/km$totss
```


#### Graphical representation of clusters
```{r}
fviz_cluster(km, data = data_kmeans)
```


#### Cluster size
```{r}
km$size
```


#### Average characteristics per cluster
```{r}
aggregate(data_kmeans, by=list(cluster=km$cluster), mean)
```

--------------------------------------------------------------------------------------------------

# Correspondence Analysis 

Here we will try to do a factorial correspondence analysis in order to bring out a risk profile by class 1 or 0.

#### Variables selected by the PCA
```{r}
data_acf <- subset(data, select = c(open_acc, total_acc, mo_sin_rcnt_tl, num_il_tl, num_rev_accts, num_tl_op_past_12m, tot_hi_cred_lim, total_il_high_credit_limit, open_acc_6m, open_il_12m, open_il_24m, open_rv_24m, inq_last_12m, loan_amnt, int_rate, installment, annual_inc, fico_range_low , revol_bal, last_fico_range_high, avg_cur_bal, total_bc_limit, max_bal_bc, mo_sin_rcnt_rev_tl_op, mths_since_recent_bc, num_actv_bc_tl, pct_tl_nvr_dlq, open_act_il, all_util, bc_util, dti, mths_since_last_delinq, mths_since_last_major_derog, mths_since_recent_bc_dlq, mths_since_recent_revol_delinq, num_accts_ever_120_pd))
```


#### ACF initialization
```{r}
res.acf <- CA(data_acf, graph = FALSE)
```


#### Graphical representation of the correspondence analysis
```{r}
fviz_ca_biplot(res.acf, map ="rowprincipal",
               repel = TRUE,
               arrow = c(FALSE, TRUE))
```


#### Eigenvalues : % of variance explained by each factorial axis
```{r}
eig.acf <- get_eigenvalue(res.acf)
```


#### Plot of inertia explained (in %) by each factorial axis
```{r}
fviz_eig(res.acf, addlabels = TRUE, ylim = c(0, 50))
```


#### Contribution of each variable to the axes
```{r}
var.contrib.col <- res.acf$col$contrib
var.contrib.col

fviz_ca_col(res.acf, col.col = "contrib",
            gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
            repel = TRUE)
```


#### Cos2 : quality of the representations
```{r}
var.cos2.cum.col <- res.acf$col$cos2[,1]+res.acf$col$cos2[,2]
var.cos2.cum.col

fviz_ca_col(res.acf, col.col = "cos2",
            gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"), 
            repel = TRUE)
```


# Analysis by class

#### Variables selected by the PCA
```{r}
data_acf <- subset(data, select = c(loan_status, open_acc, total_acc, mo_sin_rcnt_tl, num_il_tl, num_rev_accts, num_tl_op_past_12m, tot_hi_cred_lim, total_il_high_credit_limit, open_acc_6m, open_il_12m, open_il_24m, open_rv_24m, inq_last_12m, loan_amnt, int_rate, installment, annual_inc, fico_range_low , revol_bal, last_fico_range_high, avg_cur_bal, total_bc_limit, max_bal_bc, mo_sin_rcnt_rev_tl_op, mths_since_recent_bc, num_actv_bc_tl, pct_tl_nvr_dlq, open_act_il, all_util, bc_util, dti, mths_since_last_delinq, mths_since_last_major_derog, mths_since_recent_bc_dlq, mths_since_recent_revol_delinq, num_accts_ever_120_pd))
```


#### Filter on the target variable to create the two classes
```{r}
data.pos <- filter(data_acf, loan_status == 0)
data.neg <- filter(data_acf, loan_status == 1)
```


#### We delete the target variable
```{r}
data.pos <- subset(data.pos, select = -c(loan_status))
data.neg <- subset(data.neg, select = -c(loan_status))
```


#### ACF for the non-default class
```{r}
acf.neg <- CA(data.neg, graph = FALSE)

fviz_ca_biplot(acf.neg, map ="rowprincipal",
               repel = TRUE,
               arrow = c(FALSE, TRUE))
```


#### ACF for the class-default
```{r}
acf.pos <- CA(data.pos, graph = FALSE)

fviz_ca_biplot(acf.pos, map ="rowprincipal",
               repel = TRUE,
               arrow = c(FALSE, TRUE))
```

--------------------------------------------------------------------------------------------------
